---
title: "L10 Factors"
subtitle: "Data Science 1 with R (STAT 301-1)"
author: "Ryan Nguyen"

format:
  html:
    toc: true
    embed-resources: true
    code-fold: show
    link-external-newwindow: true
    
execute:
  warning: false
  
from: markdown+emoji  
---

## Overview

The goal of this lab is to learn how to deal with factors in R, specifically within the tidyverse (although many of these methods can be applied with base R functions as well). Factors are used to work with categorical variables, or variables that have a fixed and known set of possible values. We'll use the `forcats` package, which provides tools for dealing with categorical variables. 

For more information on the `forcats` package, see [`forcats` tidyverse homepage](http://forcats.tidyverse.org/reference/index.html).

## Load Packages

```{r}
#| label: load-pkgs

# Loading package(s)
library(tidyverse, quietly = T)

```

## Datasets 

We will be using the `gss_cat` dataset that is included in the `forcats` package. To view the documentation for the dataset, use `?gss_cat`.

## Exercises

### Exercise 1 

Create a bar chart to explore the distribution of `rincome` (reported income). What makes the default bar chart hard to understand? Improve the bar chart.

```{r}
income_lvl <- c("Lt $1000", "$1000 to 2999", "$3000 to 3999", 
                "$4000 to 4999", "$5000 to 5999", "$6000 to 6999", 
                "$7000 to 7999", "$8000 to 9999", "$10000 - 14999", 
                "$15000 - 19999", "$20000 - 24999", "$25000 or more",
                "Don't know", "No answer", "Not applicable", "Refused")

gss_clean <- gss_cat %>% 
  mutate(rincome = factor(rincome,
                          levels = income_lvl))

ggplot(gss_clean, aes(x = rincome))+
  geom_bar() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

**ANSWER:**
There area few issue with the bar chart. One, the labels on the x-axis are hard to read. Additionally, it does not appear sequentially and the answers that include `refused`, `don't know`, and `no answer` are mixed in with the numbers, so we want to make sure we reorder them so we can read the visualization. 

### Exercise 2

What is the most common `relig` in this survey? What’s the most common `partyid`?

```{r}
gss_cat %>% 
  group_by(relig) %>% 
  count() %>% 
  arrange(desc(n))

gss_cat %>% 
  group_by(partyid) %>% 
  count() %>% 
  arrange(desc(n))
```

**ANSWER:**
The most common `relig` in this survey is Protestant, and the most common `partyid` is Independent.

### Exercise 3

Which `relig` does `denom` (denomination) apply to? Create a visualization to find out.

```{r}
ggplot(gss_cat, aes(x = relig, fill = denom)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
**ANSWER:**
According to the visualization, `denom` mainly applies to Protestant, and also Christian. 

### Exercise 4

There are some suspiciously high numbers in `avg_tvhours` (mean or average of `tvhours`). Since the mean is not robust to outliers, it is not a good summary of this variable. Create a graphic similar to the one below, but use a more appropriate summary of `tvhours`.  

```{r}
#| label: ex-04

# data wrangling
relig_summary <- gss_cat %>%
  group_by(relig) %>%
  summarise(
    age = median(age, na.rm = TRUE),
    avg_tvhours = median(tvhours, na.rm = TRUE),
    n = n()
  )

# bar plot
relig_summary %>%
  mutate(relig = fct_reorder(relig, avg_tvhours)) %>%
  ggplot(aes(avg_tvhours, relig)) +
    geom_point()
```

### Exercise 5

How have the proportions of people who identify as Democrat, Republican, and Independent changed over the time period from 2000 to 2014? Build a visualization to display this information. You should also write a few sentences to summarize the trends within the graphic.

```{r}
gss_party <- gss_cat %>%
  mutate(political = fct_collapse(partyid,
                                other = c("No answer", "Don't know", "Other party"),
                                rep = c("Strong republican", "Not str republican"),
                                ind = c("Ind,near rep", "Independent", "Ind,near dem"),
                                dem = c("Not str democrat", "Strong democrat"))) %>% 
  group_by(year) %>% 
  mutate(count = n()) %>% 
  group_by(year, political) %>% 
  mutate(prop = n()/count)

gss_party %>% 
  ggplot(aes(year, prop, color = political)) +
  geom_line() + 
  scale_x_continuous(breaks = c(2000:2014)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
```

### Exercise 6

Demonstrate how to collapse `rincome` into a smaller set of categories.

```{r}

gss_income <- gss_cat %>%
  mutate(incomes = fct_collapse(rincome,
                                other = c("No answer", "Don't know", "Refused", "Not applicable"),
                                lower = c("Lt $1000", "$1000 to 2999", "$3000 to 3999", "$4000 to 4999"),
                                middle = c("$5000 to 5999", "$6000 to 6999", "$7000 to 7999", "$8000 to 9999"),
                                high = c("$10000 - 14999","$15000 - 19999", "$20000 - 24999", "$25000 or more")))

gss_income
```
